A comparison of model-based and design-based approaches for spatial data.

### 2021-06-08 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
introduction revision.

@@

There are two approaches to statistical inference: design-based and model-based. Both approaches aim to study a population. Because data generally cannot be sampled at all units in the population (known as population units), the population units must be sampled. In the design-based approach, the samples are often used to estimate parameters of a population. Properties of these estimators are derived through random selection of the samples. In the model-based approach, the samples are often used to predict random variables of a population. Properties of these predictions are derived through a set of distributional assumptions that specify the population's structure. Randomness is crucial to both approaches but for different reasons; the design-based approach uses randomness when selecting a sample of population units, and the model-based approach uses randomness when specifying distributional assumptions. @sterba2009alternative provides historical context for each approach, and @hansen1983evaluation discusses the advantages of each approach [@hansen1983evaluation].

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus comparing these approaches for spatial data. @de1990model gave an early comparison of these approaches, quashing the belief that design-based approaches could not be used for spatially correlated data. Thereafter, several comparisons between design-based and model-based for spatial data have been considered, but they tend to compare design-based approaches that ignore spatial location in sampling to model-based approaches [@brus1997random, @verhoef2002sampling, @verhoef2008spatial]. No comparison has been made design-based approaches that incorporate spatial location in sampling and model-based approaches.

define sampling design, define parameters, define model-based, sample, population, super-population?, etc.

introduction of spatial data / definition of spatial and maybe general background?

give some examples of possible spatial data

2-3 sentences about each design

goals and outline of paper

brus2020statistical: most similar to what this paper will be, directly comparing the two approaches. This has a lot of other potentially useful references to go back to. (published in 2021 so likely doesn't have many that have cited it yet).

wang2012spatial - sampling and inference steps, thorough overview in terms of spatial context, lots of references

@@

In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach. In Section \ref{sec:numstudy}, we use simulated and real data to compare the properties of parameter estimates from both aproaches. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

>>>>

finish spatial introduction
revise section 2
reread brus and wang
write down potential data sets

@@

There are two general approaches for using data to make statistical inferences about a population: design-based approaches and model-based approaches. When data cannot be obtained for all units in a population (known as population units), data on a subset of the population units is collected in a  sample. In the design-based approach, inferences about the underlying population are informed from a probabilistic process in which population units are selected to be in the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions made about the underlying process that generated the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of general advantages [@hansen1983evaluation].

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data.

<!-- ????????? -->
We define spatial data as [blah blah blah, any data collected at specific locations blah blah blah blah]. Examples include [do we want to give a list of general examples?]. 
<!-- ???????????? -->

@de1990model give an early comparison of these approaches, quashing the belief that design-based approaches could not be used for spatially correlated data. Thereafter, several comparisons between design-based and model-based for spatial data have been considered, but they tend to compare design-based approaches that ignore spatial location in sampling to model-based approaches [@brus1997random; @verhoef2002sampling; @verhoef2008spatial]. More recent overviews include @brus2020statistical and @wang2012spatial, but no numerical comparison has been made between design-based approaches that incorporate spatial location in sampling and model-based approaches.

@@

The rest of this paper is organized as follows. In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach. In Section \ref{sec:numstudy}, we use simulated and real data to compare the properties of parameter estimates from both aproaches. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

## Applied Example

### 2021-06-08 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
introduction revision.

% Pandoc citation processing

@@

There are two approaches to statistical inference for spatial data:
design-based and model-based. In the design-based approach, statistical
inference is based on random sampling. In the model-based approach,
statistical inference is based on a set of distributional assumptions
for the data.

In either approach, statistical inference requires estimating parameters
of interest. Often, these parameters are means or totals. In the
design-based approach, parameters correspond to some population. In the
model-based approach, parameters correspond to some statistical process,
sometimes called a superpopulation.

Spatial data paragraph

define sampling design, define parameters, define model-based, sample,
population, super-population?, etc.

introduction of spatial data / definition of spatial and maybe general
background?

give some examples of possible spatial data

2-3 sentences about each design

goals and outline of paper

De Gruijter and Ter Braak (1990) give an early comparison of the
paradigms for spatial data, quashing the belief that classical
sampling-based methods could not be used for spatially correlated data.
Thereafter, several comparisons between model-based and design-based for
spatial data have been considered, but they tend to compare spatial
models to IRS (Independent Random Sampling) or SIRS (stratified IRS)
(Brus and De Gruijter, 1997). No comparison has been made between
spatial models and spatially balanced design-based approaches.

brus2020statistical: most similar to what this paper will be, directly
comparing the two approaches. This has a lot of other potentially useful
references to go back to. (published in 2021 so likely doesn't have many
that have cited it yet).

\begin{itemize}
\item
  hansen1983evaluation - general overview (in non spatial context), talk
  about population vs superpopulation give a set of guiding principles
  in the discussion
\item
  verhoef2002sampling - strictly defines classical sampling, compares
  FPBK to classical, BK to classical
\item
  wang2012spatial - sampling and inference steps, thorough overview in
  terms of spatial context, lots of references
\end{itemize}

In Section \ref{sec:background}

@@

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g.~Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\hypertarget{numerical-study}{%
\section{Numerical Study}\label{numerical-study}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.

@@

@@

\leavevmode\hypertarget{ref-chiles1999geostatistics}{}%
Chiles, J.-P., Delfiner, P., 1999. Geostatistics: Modeling Spatial
Uncertainty. John Wiley \& Sons, New York.

@@

@@

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-nn) methods for forestry applications. PloS
one 8, e59129.

@@

>>>>

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

There are two general approaches for using data to make statistical
inferences about a population: design-based approaches and model-based
approaches. When data cannot be obtained for all units in a population
(known as population units), data on a subset of the population units is
collected in a sample. In the design-based approach, inferences about
the underlying population are informed from a probabilistic process in
which population units are selected to be in the sample. Alternatively,
in the model-based approach, inferences are made from specific
assumptions made about the underlying process that generated the data.
Each paradigm has a deep historical context (Sterba, 2009) and its own
set of general advantages (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data.

We define spatial data as {[}blah blah blah, any data collected at
specific locations blah blah blah blah{]}. Examples include {[}do we
want to give a list of general examples?{]}.

De Gruijter and Ter Braak (1990) give an early comparison of these
approaches, quashing the belief that design-based approaches could not
be used for spatially correlated data. Thereafter, several comparisons
between design-based and model-based for spatial data have been
considered, but they tend to compare design-based approaches that ignore
spatial location in sampling to model-based approaches (Brus and De
Gruijter, 1997; Brus, 2020; Ver Hoef, 2002; Ver Hoef, 2008). No
comparison has been made between design-based approaches that
incorporate spatial location in sampling and model-based approaches.

wang2012spatial - sampling and inference steps, thorough overview in
terms of spatial context, lots of references

In Section \ref{sec:background}, we compare sampling and estimation
procedures between the design-based approach and the model-based
approach. In Section \ref{sec:numstudy}, we use simulated and real data
to compare the properties of parameter estimates from both aproaches.
And in Section \ref{sec:discussion}, we end with a discussion and
provide directions for future research.

@@

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g. Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\hypertarget{sec:numstudy}{%
\section{Numerical Study}\label{sec:numstudy}}

@@

\hypertarget{sec:discussion}{%
\section{Discussion}\label{sec:discussion}}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.

@@

\leavevmode\hypertarget{ref-brus2020statistical}{}%
Brus, D.J., 2020. Statistical approaches for spatial sample survey:
Persistent misconceptions and new developments. European Journal of Soil
Science.

@@

\leavevmode\hypertarget{ref-chiles1999geostatistics}{}%
Chiles, J.-P., Delfiner, P., 1999. Geostatistics: {Modeling Spatial
Uncertainty}. {John Wiley \& Sons}, New York.

@@

\leavevmode\hypertarget{ref-hansen1983evaluation}{}%
Hansen, M.H., Madow, W.G., Tepping, B.J., 1983. An evaluation of
model-dependent and probability-sampling inferences in sample surveys.
Journal of the American Statistical Association 78, 776--793.

@@

\leavevmode\hypertarget{ref-verhoef2002sampling}{}%
Ver Hoef, J., 2002. Sampling and geostatistics for spatial data.
Ecoscience 9, 152--161.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-NN) methods for forestry applications. PloS
one 8, e59129.

@@

\end{CSLReferences}
